---
title: "Homework 2 - Farouk Alghazzy"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=T,eval=T,message=F,warning=F,fig.align='center')
library(tidyverse)
```

## 1. Independence of Events <small>(8 points, 2 points each)</small>

A fair four-sided die (numbered 1,2,3 and 4) is rolled three times. Consider the following four events:

A: "The sum is 6 or less"

B: "The second roll is more than the first"

C: "The sum is odd"

D: "Exactly two of the rolls are the same"
```{r}
q = expand.grid(1:4, 1:4, 1:4)

qa = q %>%
  mutate(sum = Var1 + Var2 + Var3) %>%
  filter(sum <= 6) %>%
  count()
  
qa_prob = qa$n/64

qb = q %>%
  mutate(sum = Var1 + Var2 + Var3) %>%
  filter(Var1 > Var2) %>%
  count()

qb_prob = qb$n/64

qc = q %>%
  mutate(sum = Var1 + Var2 + Var3) %>%
  filter(sum == 3 | sum == 5 | sum == 7 | sum == 9 | sum == 11) %>%
  count()
  
qc_prob = qc$n/64

qd = q %>%
  mutate(sum = Var1 + Var2 + Var3) %>%
  filter(Var1 == Var2 | Var1 == Var3 | Var2 == Var3) %>%
  filter(!(Var1 == Var2 & Var2 == Var3)) %>%
  count() 
  
qd_prob = qd$n/64
```
a. Are A and B independent?
```{r}
qa1 = q %>%
  mutate(sum = Var1 + Var2 + Var3) %>%
  filter(sum <= 6) %>%
  filter(Var1 > Var2) %>%
  count()

qa1_prob_a = qa1$n/64
qa1_prob_a

qa1_prob_b = qa_prob * qb_prob
qa1_prob_b
#A and B are not independent because they have different probabilities.
```

b. Are C and D independent?
```{r}
qb1 = q %>%
  mutate(sum = Var1 + Var2 + Var3) %>%
  filter(sum == 3 | sum == 5 | sum == 7 | sum == 9 | sum == 11) %>%
  filter(Var1 == Var2 | Var1 == Var3 | Var2 == Var3) %>%
  filter(!(Var1 == Var2 & Var2 == Var3)) %>%
  count() 

qb1_prob_a = qb1$n/64
qb1_prob_a

qb1_prob_b = qc_prob * qd_prob
qb1_prob_b
# C and D are independent, because their probabilities are the same.
```

c. Are B and C independent?
```{r}
qb = q %>%
  filter(Var2 > Var1) %>%
  count()
qb_prob = qb$n / 64
qc = q %>%
  mutate(sum = Var1 + Var2 + Var3) %>%
  filter(sum %% 2 == 1) %>%
  count()
qc_prob = qc$n / 64
qc1 = q %>%
  mutate(sum = Var1 + Var2 + Var3) %>%
  filter(Var2 > Var1) %>%
  filter(sum %% 2 == 1) %>%
  count()

qc1_prob_a = qc1$n / 64
qc1_prob_b = qb_prob * qc_prob

qc1_prob_a
qc1_prob_b
#B and C are independent because they share the same individual probabilities.
```

d. Are B and D independent?
```{r}
qd1 = q %>%
  mutate(sum = Var1 + Var2 + Var3) %>%
  filter(Var1 > Var2) %>%
  filter(Var1 == Var2 | Var1 == Var3 | Var2 == Var3) %>%
  filter(!(Var1 == Var2 & Var2 == Var3)) %>%
  count()

qd1_prob_a = qd1$n/64
qd1_prob_a

qd1_prob_b = qb_prob * qd_prob
qd1_prob_b
#B and D are not independent because they don't share the same probabilities.
```


## 2. Stock Portfolio <small>(8 points, 2 points each)</small>

Consider two stocks. Stock A has a mean annual return of 0.07 and a standard deviation of 0.015. Stock B has a mean annual return of 0.15 and a standard deviation of 0.07. Past data suggests that the prices have a correlation of -0.75.
You have \$1000 to invest in a portfolio composed of these two stocks. 

a. Suppose you invest \$500 in stock A and \$500 in stock B. What will be the expected return and the standard deviation of your portfolio (in dollars)? 
*Hint:* $\text{SD}(aX+bY) = \sqrt{\text{Var}(aX+bY)}= \sqrt{a^2\text{Var}(X)+2ab\text{Cov}(X,Y)+b^2\text{Var}(Y)}$ and $\text{Cov}(X,Y)=\sigma_X \sigma_Y \rho_{XY}$
```{r}
return1 = .07 * 500
return2 = .15 * 500
x = return1 + return2
x
```
```{r}
cov = (-.75)*(.015*.07)
return_std = sqrt((500*500)*(0.015*0.015) + (500*500)*(0.07*0.07) + 2*500*500*cov)
return_std
```


b. Give an expression for the standard deviation for a portfolio investing $\$x$ in stock A and $\$1000-x$ in stock B. 
```{r}
std = sqrt(x^2)*(.015^2)+ ((1000-x)*(1000-x)*(0.07^2)) + ((2)*(x)*(1000-x)*cov)
```

c. Find the value $x$ that minimizes this standard deviation. 
```{r}
minimize = function(x) {
  sqrt(x^2 * .015^2 + (1000-x) * .07^2 + (2 * x * (1000-x) * (-.75 * .07*.015)))
}
minimizer = optimize(minimize, interval = c(0,1000))[1]
minimizer
```

d. What is the expected return and standard deviation (in dollars) for the portfolio split you chose in part c?
```{r}
return_a = 0.07 * 874.6097
return_b = 0.015 * (1000-874.6097)
return_ab = return_a + return_b
return_ab
```

```{r}
#STD
x = 874.60971
cov = (-0.75) * (0.015 * 0.07)
return_std = sqrt(x^2 * 0.015 * 0.015 + (1000-x)^2 * 0.07 * 0.07 + (2 * x * (1000-x) * cov))
return_std
```


## 3. Mousetrap: A Rube Goldberg Machine  <small>(5 points, 1 points each)</small>

Consider the board game [Mousetrap](https://youtu.be/fOyq6Nc8660?si=WFJ-ldFsqsxg0Khx). When you trigger the trap, the ball needs to 

1. roll down a **ramp**,
2. be carried up by a **hammer**,
3. roll  down another **ramp**,
4. go down a bathtub **drain**,
5. land on a **lever** which
6. sends a diver into a **basin** which then
7. causes the **trap** to descend.

Suppose you can model the success of each component as independent Bernoulli random variables $X_1$ through $X_7$, where the probability of $X_i$ succeeding is $.90$ for each component (this is an arbitrary probability for this HW problem). 

a. What is the probability that the trap works correctly from beginning to end?
```{r}
execution_prob = (0.9)^7
execution_prob
```

b. If you attempt the trap 15 times during a game, how many times do you expect it to work correctly from beginning to end?
```{r}
execution_x = execution_prob * 15
execution_x
```

c. Suppose you let the marble go: Let $Y$ be the number of components of the trap that work in a row. If one component fails, then you stop counting. Calculate the probability mass function of $Y$. *Hints: $Y$ can have values $0, 1, 2, 3, 4, 5, 6$ or $7$. This is like a Geometric r.v. except it does not have infinite support. How can you use this fact to determine the pmf? Be careful: what is a 'success' in this context?*
```{r}
# P(Y = 0) = 0.1
# P(Y = 1) = (0.9) * 0.1 or 0.09
# P(Y = 2) = (0.9^2) * 0.1 or 0.081
# P(Y = 3) = (0.9^3) * 0.1 or 0.0729
# P(Y = 4) = (0.9^4) * 0.1 or 0.06561
# P(Y = 5) = (0.9^5) * 0.1 or 0.05905
# P(Y = 6) = (0.9^6) * 0.1 or 0.05314
# P(Y = 7) = (0.9^7) or 0.4783
trap_prob = c(0.1, 0.09, 0.081, 0.0729, 0.06561, 0.05905, 0.05314, 0.4783)
trap_prob
```


d. Calculate the expected value of $Y$ and its variance. 
```{r}
exp_trap = sum(trap_prob * 0:7)
exp_trap
```
```{r}
var = sum(trap_prob * (0:7)^2) - exp_trap^2
var
```

e. What is $P(Y = 6 | Y \geq 3)$?
```{r}
x = trap_prob[7]
y = sum(trap_prob[4:8])

prob = x / y
prob
```


## 4. Bayes Rule with backyard animals <small>(9 points, 3 points each)</small>

In my backyard I sometimes see a black squirrel and sometimes a hawk. I see the squirrel roughly 2/3 of days during the year (equal probability year round). I see a hawk 85% of days when the squirrel is present, and 40% of days when the squirrel is not present.

a. Calculate the overall probability that I see only one of the two animals on a given day.
```{r}
P_S = 2/3
x = 0.85
y = 0.4

P_S_andnotH = P_S * (1 - x)
P_notS_and_H = (1 - P_S) * y
P_only_one = P_S_andnotH + P_notS_and_H
cat("Prob of seeing only one animal:", P_only_one)
```

b. What is the expected number of days in a year that I would not see either animal?
```{r}
P_S = 2/3
x = 0.85
y = 0.4
P_neither = (1- P_S) * (1-y)
days = 365
expected_days = P_neither * days
cat("Expected number of days without seeing either animal:", expected_days)

```

c. I post a really nice photo of the hawk to my social media. What is the probability that the squirrel also visits this day?

```{r}
P_S = 2/3
x = 0.85
y = 0.4
P_H = (x * P_S) + (y *(1 - P_S))
P_S_GH = (x * P_S) / P_H
cat("Prob that the squirrel visits given that the hawk is seen:", P_S_GH)
```

